Network Data Model and BerlinMOD Benchmark

### 2010-07-05 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Started section implementation issues

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.x}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, and the translation
from the \bmodb{} into the network constrained data model and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space and data models
for network constrained objects. For both categories several diffferent data models
have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis,HERMESMDCPelekis}
for spatio-temporal data objects moving freely in two dimensional space (DMFS) and
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,NetworkVazirgiannis}
for spatio-temporal data objects that are constrained by given networks (NCDM),
to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

Why do we spend time on NCDM,if everything can be represented by
a DMFS? Now, it is natural to give positions related to the street
network instead of x,y-coordinates. NCDM are expected to use less storage space,
because geographical informations about street curves are stored only once in
the network, whereas in DMFS each street curve is stored in each moving point
object using this street. NCDM can support query processing with spezialized indexes,
because of the knowledge of the underlying network. And it is much easier to
formulate queries about the relationships between moving objects and the network
in the NCDM. And last but not least the results of our experiments show that
our network data model outperforms our data model of free movement in two dimensional
space by orders of magnitude. The network constrained data model uses less than
60\% of the storage space and less than 50\% of the total query run time of the
data model of free movement in space we used in our experiments.
We think that these results show that it is useful
to develop specialised data models for specialized data structures like NCDM for
newtork constrained moving objects to save storage space and reduce query run times.

For our benchmark experiments, presented in this paper, we chosed two data models
one for each data model category. Both data models use the same temporal representation
and are available in \secondo{} DBMS
\cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting}.
So we can exclude that different DBMS or temporal
representation issues bias the results of our data model comparison with the
\bmodb{}\cite{BerlinMODVLDBDuentgen}.
The DMFS we use is the data model presented in \cite{DataModelDataStructureGueting}
(SPACE for short).
And the NCDM we use is the data model presented in \cite{NetworkGueting} (NET for short).
Our implementation of NET is described first in this paper (Section \ref{sec:implNDM}
and provides some further concepts which were only sketched in \cite{NetworkGueting}.

We used the \bmodb{} \cite{BerlinMODVLDBDuentgen} to compare the capabilities of
the two data models, because the \bmodb{} is to the best of our knowledge the
first benchmark for complete spatio-temporal database systems.
It is developed and available in \secondo{} DBMS. And the data generated by the
\bmodb{} data generator are restricted to the streets of the German capital Berlin,
such that they can be translated into a network constrained environment.
And last but not least, the data model used in the \bmodb{} is SPACE that we use
for our comparison. So we only have to translate the spatial and spatio-temporal
data types of the \bmodb{} once into our NET representation. This simplifies the
control of the query results and avoids errors caused by translation.
The translation of the spatial and spatio-temporal data types of the \bmodb{} data
into the NET representation described in this paper \ref{sec:Translation} can be
seen as an example for the usage of the \bmodb{} with other compatible data
representations or DBMS.

The rest of the paper is organised as follows: Related Work is presented in Section
\ref{sec:relWork}. In the sections \ref{sec:secondo} - \ref{sec:bmodb} we give
short reviews of the underlying \secondo{} DBMS (Section \ref{sec:secondo}), the two data
models (SPACE Section \ref{sec:bmodbdatamod} and NET Section \ref{sec:netdatamod})
we choose for our comparison, and the \bmodb{} (Section \ref{sec:bmodb}).
In Section \ref{sec:implNDM} we give some information about our NET implementation,
the used operations and indexes.
The translation of the \bmodb{} data and query set into the NET representation
is described in Section \ref{sec:Translation}. The resulting experimental benchmark
setup is described in Section \ref{sec:scenario} followed by the results of our
experiments in Section \ref{sec:results}. We conclude our work in Section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past many different spatio-temporal data models have been presented. Many
of them support only discrete spatio-temporal changes
\cite{sqlstchen,Ramadrachan,Langran1,Langran2,HunterWilliamson}
or deal only with current and future positions of continuously moving objects
\cite{MOSTWolfson}. In this paper we will focus on spatio-temporal data models
for complete histories of continuously moving objects.

To the best of our knowledge only a few of these data models have been completely
implemented into different database management systems:
\begin{enumerate}
  \item HERMES \cite{HERMESMDCPelekis, HERMESPelekis} is implemented as data
cartridge for the commercial Oracle\textregistered ORDBMS \cite{Oracle}. The
used data model is the DMFS STAU \cite{STAUPelekis}.
  \item The Dynamic Transportation Network \cite{DynamicTransportNetworkDing} is
a NCDM implemented in the open source database project PostgreSQL \cite{PostgreSQL}.
  \item In the freely available \secondo{} DBMS \cite{SecondoPlatformPrototypingGueting}
two spatio-temporal data models have been implemented.
\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
as DMFS and \cite{NetworkGueting} as NCDM.
\end{enumerate}

Because \secondo{} provides a data model for each of our catgories we use \secondo{}
and the provided data models. Although the data model of HERMES provides more
capabilities than SPACE.
Both data models use the basic idea of a sliced representation for moving objects
introduced by \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}.
The implementation of the sliced representation in SPACE supports only linear
interpolation, whereas HERMES arc interpolation is also supported. Both data
models support time varying geometries, which change their positions and / or
extent in space and time dimension. HERMES supports discret and continuous
changes, whereas SPACE only supports continuously moving objects. HERMES uses
one spatial object for different spatial geometries, whereas SPACE
uses different spatial objects for the different spatial geometries.
The same holds for operations on topological relationships. SPACE
uses different operations, whereas HERMES only provides a single operator which
distinguishes between the different topological relationships via a ``mask'' parameter.

Although HERMES provides a greater flexibility we decided to use SPACE in our
experiments, because the \secondo{} DBMS provides besides
the SPACE data model a NCDM \cite{NetworkGueting} (see Section \ref{sec:netdatamod}),
and with the \bmodb{} \cite{BMODBVLDBDuentgen} (see \ref{sec:bmodb})
to the best of our knowledge the first benchmark for complete spatio-temporal
database systems.

Other existing benchmarks \cite{COSTBenchmarkJensen,QueriesTheodoridis}

like \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting,STAUPelekis,HERMESMDCPelekis}
and data models dealing with network constrained moving objects (NCDM) like \cite{NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}.

and \cite{NetworkGueting} as NCDM. Therefore we describe these data models
in the sections \ref{sec:bmodbdatamod} - \ref{sec:netdatamod} in more detail.

\cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting,NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing},As mentioned before we can categorize spatio-temporal data models into
data models dealing with current and future positions of moving objects,
data models dealing with complete histories of moving objects, and data models
dealing with past, current and future positions of moving objects. We will focus
on data models dealing with complete histories of moving objects, like \cite{STAUPelekis,HERMESMDCPelekis,RepresentingMovingObjectsGueting,DataModelDataStructureGueting,NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}, to name just a few.

DMFS\cite{STAUPelekis,HERMESPelekis,RepresentingMovingObjectsGueting,DataModelDataStructureGueting,sqlstchen}

NCDM\cite{NetworkGueting,NetworkJensen,NetworkVazirgiannis,DynamicTransportationNetworkDing}.

Also, there are some more
benchmarks \cite{COSTBenchmarkJensen,QueriesTheodoridis} and database systems for
spatio-temporal data types \cite{SecondoPlatformPrototypingGueting,HERMESPelekis}. But to the best of our
knowledge they do not provide a combination of different implemented
and supported data models together with an existing benchmark feasible for moving
objects in space and network constrained moving objects like the current \secondo{} DBMS.
So it is natural for us to use the \secondo{} DBMS in combination with the provided
data models and the \bmodb{} to compare the capabilities of the two data models.

In the next sections we give short reviews of the \secondo{} database system
(Section \ref{sec:secondo}), the both data models (Section \ref{sec:bmodbdatamod},
Section \ref{sec:netdatamod}) we used in our experiments, and the \bmodb{}
(Section \ref{sec:bmodb}).
\section{\secondo{} DBMS}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce new data types and operations on
these data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the users may publish their extensions
as a \secondo{} plugin such that other users can use this plugin to extend their own
\secondo{} system. They may use the newly provided functionalities or repeat the
published experiments. \secondo{} is freely available on the web \cite{secondoweb}.
It comes with a number of already implemented spatial and spatio-temporal data types
and operations including the DMFS (Section \ref{sec:bmodbdatamod}) and the NDM
(Section \ref{sec:netdatamod}) used in this paper. Furthermore, the \bmodb{}
described in Section \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version \secver{}.
\section{Data Model of Free Movement in Two Dimensional Space(SPACE)}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the DMFS presented in \cite{STDTModelingQueryingMOinDErwig,DataModelDataStructureGueting,RepresentingMovingObjectsGueting}.
Spatial positions are assumed to be located in a two dimensional space.

A single spatial
position is represented by the data type \dt{point}. A \dt{point} consists of a pair of
\dt{real} values interpreted as x,y-coordinates in the assumed two dimensional plane.

Streets are represented by \dt{line} values. A \dt{line} value consists of a set
of \dt{HalfSegment}s representing the geometry of the line in the two dimensional space.
Each \dt{HalfSegment} consists of two \dt{point} values and a Boolean flag telling us
if the left or the right point is the dominating point of the \dt{HalfSegment}.

Regions are represented by the data type \dt{region}. A \dt{region} consists of
a set of \dt{HalfSegment}s defining the outer border of the region in
the two dimensional space. If a region contains wholes the inner border is also
formed by the \dt{HalfSegment}s.

In \secondo{} all these spatial data types and many standard data types can be
''lifted'' to become time dependent \dt{moving} values. For all data types \dt{$\alpha$}
the constructor \dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$})
(short form \dt{m$\alpha$}).

A car may be represented by a \dt{moving}(\dt{point}), short \dt{mpoint}.
An \dt{mpoint} consists of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first \dt{point} value represents the
position of the \dt{mpoint} at the start of the time interval and the second
\dt{point} value represents the position of the \dt{mpoint} at the end of the
time interval. It is assumed that the object represented by the \dt{mpoint}
moves on the straight line between these two points with constant speed within the
given time interval. The velocity of the object is given by the ratio from the
distance of the two points and the length of the time interval of the unit.
All units of an \dt{mpoint} must have disjoint time intervals, because a car
cannot be at two different positions at the same time.
The units are sorted by ascending time intervals.

This spatio-temporal data model of \dt{moving} allows us to compute the position
of an \dt{mpoint} at every time instant within its definition time.
We can also compute the time instant the point passed a
given position assuming the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by an \dt{intime}(\dt{point})
(short form \dt{ipoint}). An \dt{ipoint} consists of a time instant and a \dt{point} value.

Some other data types of \secondo{} which are used in the \bmodb{} are shown in
Table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{1.0\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type\\
\hline
\dt{int} & Usual integer number\\
\hline
\dt{real} & Usual real number\\
\hline
\dt{instant} & A point in time\\
\hline
\dt{periods} & A set of disjoint and not connected time interval\\
\hline
\dt{mbool} & A time dependent boolean value, which is constant \true{} or \false{}
within each \dt{ubool} \\
\hline
\dt{mreal} & Time dependent real number. Each unit is defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\section{Network Data Model(NET)}
\label{sec:netdatamod}
The central idea of the NDM presented in \cite{NetworkGueting} is that
every movement is constrained by a given network and every position can be described
relative to this network. The data type \dt{network} is the central
data type in the NDM. All other data types of the NDM
are related to a given \dt{network} object by the unique network identifier that
is part of each \dt{network} object.

The \dt{network} object contains all spatial information of the represented network.
The \dt{network} consists of three main relations ($routes$, $junctions$, and $sections$),
two arrays providing fast access to adjacent network sections,
and some B-Tree and R-Tree indexes to supporting faster access to the main relations.

The relation $routes$ of the \dt{network} contains the attributes of the
streets like $id$, $route curve$, $route length$, and two Boolean flags. The first
flag indicates if the route starts at the lexicographically smaller end point or not.
The second flag indicates if the lanes of the street are separated like on German
Highways or not.

The $junctions$ relation of the \dt{network} contains all attributes of the street
crossings like the two route identifiers of the first and second street meeting at the
crossing, the distance of the crossing from the start of the first respectively
second street, tuple identifiers of the both streets in the routes relation,
tuple identifiers of the sections connected by this junction in the sections relation,
and a connectivity code telling us which lanes of the two streets are connected
by the crossing.

The $sections$ relation of the \dt{network} object contains the
attributes of the street parts between two crossings or a crossing and
the end of the street. These are the route identifier of the street the section
belongs to, the tuple identifier of this street in the routes relation, the start
position and end position of the section on the street, the section curve, and
again two Boolean flags with the same meaning as in the routes relation.

Furthermore there are two arrays in the \dt{network} object providing a fast
access from each section to their adjacent sections with respect to the driving
direction. Two sections are adjacent if their lanes are connected by a junction.

We created four B-Tree indexes for the route identifier attributes in the $routes$,
$junctions$ and $sections$ relation, and an R-Tree index over the curve attribute
of the $routes$ relation to support faster execution of operations dealing with
the relations of the \dt{network} object.

The data type \dt{gpoint} represents single positions in a given network. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a $side$ value
({$up$, $down$, $none$) telling us if the position is reachable from the $up$
or the $down$ side of the route in case of separated lanes. For simple
streets or positions which are reachable from both sides of the route, the side
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier, a \dt{gline} consists of a set
of \dt{RouteInterval}s, and two Boolean flags. The Boolean flags tell us if the
\dt{gline} is defined and if the set of \dt{RouteInterval}s is sorted.

>>>>

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage[utf8x]{inputenc}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[ruled]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\usepackage{todonotes}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, the translation
from the \bmodb{} into the network constrained data model, and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\listoftodos
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space (DMFS) and data models
for network constrained moving objects (NCDM). For both categories several different
data models have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis,HERMESMDCPelekis}
for DMFS and \cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,NetworkVazirgiannis}
for NCDM, to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

Why do we spend time on NCDM, if everything can be represented by DMFS?
Now, it is natural to give positions related to the street
network instead of x,y-coordinates. NCDM are expected to use less storage space,
because geographical information's about street curves are stored only once in
the network, whereas in DMFS each street curve is stored in each moving point
object using this street. NCDM can support query processing with specialized indexes
using their knowledge of the underlying network. It is much easier to
formulate queries about the relationships between moving objects and the network
in the NCDM. And not at last, the results of our experiments show that
our network constrained data model outperforms our data model of free movement in
two dimensional space by orders of magnitude. The network constrained data model
uses less than 60\% of the storage space and less than 50\% of the total query
run time of the data model of free movement in space, which we used in our experiments.
We think that these results show that it is useful to develop specialized data
models for specialized data structures like NCDM for
network constrained moving objects to save storage space and reduce query run times.

For our benchmark experiments, presented in this paper, we chose two data models
one for each data model category. Both data models use the same temporal representation
and are available in \secondo{} DBMS \cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting}.
So we can exclude that different DBMS or temporal representation issues bias the
results of our data model comparison with the \bmodb{}\cite{BerlinMODVLDBDuentgen}.
The DMFS we use is the data model presented in \cite{RepresentingMovingObjectsGueting,DataModelDataStructureGueting}
(SPACE). And the NCDM we use is the data model presented in \cite{NetworkGueting} (NET).

We used the \bmodb{} \cite{BerlinMODVLDBDuentgen} to compare the capabilities of
the two data models, because the \bmodb{} is to the best of our knowledge the
first benchmark for complete spatio-temporal database systems.
It is developed and available in \secondo{} DBMS. And the data generated by the
\bmodb{} data generator are restricted to the streets of the German capital Berlin,
such that they can be translated into a network constrained environment.
And not at last, the data model used in the \bmodb{} is SPACE that we use
for our comparison. So we only have to translate the spatial and spatio-temporal
data types of the \bmodb{} once into our NET representation. This simplifies the
control of the query results and avoids errors caused by translation.
The translation of the spatial and spatio-temporal data types of the \bmodb{} data
into the NET representation described in Section \ref{sec:Translation} can be
seen as an example for the usage of the \bmodb{} with other compatible data
representations or DBMS.

Besides the comparsion of the both data models, we describe in this paper the
first real implementation of NET (see Section \ref{sec:implNDM}) providing some
further concepts which were only sketched in \cite{NetworkGueting}.

The rest of the paper is organised as follows: We present some related work in Section
\ref{sec:relWork}, including short reviews of the underlying \secondo{} DBMS
(Section \ref{sec:secondo}), the two data models (SPACE Section \ref{sec:bmodbdatamod},
NET Section \ref{sec:netdatamod}) we chose for our comparison, and the
\bmodb{} (Section \ref{sec:bmodb}).
In Section \ref{sec:implNDM} we give some information's about our implementation
of NET, the used operations and indexes.
The translation of the \bmodb{} data and query set into the NET representation
is described in Section \ref{sec:Translation}. The resulting experimental benchmark
setup is described in Section \ref{sec:scenario} followed by the results of our
experiments in Section \ref{sec:results}. We conclude our work in Section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past many different spatio-temporal data models have been presented. Many
of them support only discrete spatio-temporal changes like
\cite{sqlstchen,HunterWilliamson,Langran2,Langran1,Ramadrachan}
or deal only with current and future positions of continuously moving objects like
\cite{MOSTWolfson}. More detailed reviews of these and other spatio-temporal
data models beyond the scope of our paper can be found in \cite{ReviewSTDMPelekis}.

In this paper we will focus on spatio-temporal data models for complete histories
of continuously moving objects. These can be categorized into data models for
objects moving freely in two dimensional space (DMFS) and data models for network
constrained moving objects (NCDM).

\cite{MOLRaQSu} proposes an incomplete abstract DMFS. Basic idea is that
spatio-temporal data types can be modeled by linear constraints and queries are
formulated using formulas from differential geometry.

\cite{RepresentingMovingObjectsGueting} proposes an abstract DMFS including the
idea of an time sliced representation for moving objects. This basic idea of
time sliced representation is used by the DMFS \cite{STAUPelekis} and
\cite{DataModelDataStructureGueting}. The last one is used in our experiments
and therefore reviewed in more detail in Section \ref{sec:bmodbdatamod}.
The main difference between the both data models is that
\cite{DataModelDataStructureGueting} supports only linear interpolation of movement,
whereas \cite{STAUPelekis} also supports arc interpolation of movement.
\cite{STAUPelekis} uses only one spatial object
containing all spatial geometries for the representation of spatial objects and
one moving object for the representation of the different moving object data types,
whereas \cite{DataModelDataStructureGueting} uses different spatial and moving
objects for the representation of the different spatial and moving data types.
According to this \cite{STAUPelekis} provides only a single operator that distinguishes
between the different topological relationships via a parameter, whereas
\cite{DataModelDataStructureGueting} uses different operations to estimate
topological relationships. Overall, \cite{STAUPelekis} offers a more flexible
object oriented design than \cite{DataModelDataStructureGueting}.

The spatio-temporal framework of \cite{RepresentingMovingObjectsGueting} used by
\cite{DataModelDataStructureGueting} has been used for the definition of a
NCDM in \cite{NetworkGueting}.

The most NCDM use edge based graph representations for the representation of the
underlying network data only a small number of NCDM uses route oriented data models
or combine route and edge based data models.
.
\cite{NetworkVazirgiannis} proposes an edge based NCDM. The edges and their attributes
are stored in a relation representing the network as an undirected graph. Moving
objects are assumed to drive always on the path with the lowest cost,
in terms of distance or travel-time. They are defined by the source point, the
target point, and the starting time instant of the trip. The trajectory is
computed by this assumption using the length and speed attributes of the graph
edges within the shortest respectively fastest path computation.
The advantage of this definition is the reduced storage space for the representation
of moving point objects. The drawback is the high computational effort for query
evaluation on moving objects.

\cite{NetworkJensen} uses also an edge based network representation. The paper
proposes a combination of an two dimensional geometrical edge representation with
an directed graph representation of the same network. The both representations
are connected by transition policies. The two dimensional
geometrical representation handles the spatial information's, whereas the connectivity
information is mostly embedded in the directed graph representation.
Moving objects are represented by sets of five tuples. Each five tuple contains
an edge identifier, the position of the moving point on the edge
in terms of weight and length, the speed and direction of the movement, and
the time instant of this information.

Another two-layer network representation is proposed by \cite{DynamicTransportNetworkDing}.
The authors of \cite{DynamicTransportNetworkDing} combine the advantages of the
dynamic edge-based \cite{DynNetworkModEdgeGueting} and the dynamic route-based
\cite{DynNetworkModRouteGueting} NCDM approaches. The route-based environment
reduces the update intervals and used storage space for the database representation
of moving point objects, whereas the edge-based environment supports
a more detailed view on the traffic conditions of the different edges belonging
to the same route. Moving objects are represented by a set of pairs. The pairs
consist of an motion vector and an Boolean flag. The Boolean flag tells if the
motion vector contains current or historical information. Each motion vector
consists, in parts similar to \cite{NetworkJensen}, of a time stamp, a network
position, and a speed vector. Similar to \cite{NetworkGueting} the network
position bases on routes and junctions and not on edges. Different from all
other NCDM in this section \cite{DynamicTransportNetworkDing}
uses time depending dynamic attributes in the representation of the network parts.
Therefore, changes in the network environment can be handled without loss of
information in this NCDM.

To the best of our knowledge only a few of the proposed data models have been
implemented into database management systems: \cite{STAUPelekis} is implemented
as data cartridge \cite{HERMESMDCPelekis, HERMESPelekis} for the commercial
Oracle\textregistered object-relational DBMS \cite{oracle};
\cite{DynamicTransportNetworkDing} is implemented as extension of the open source
database project PostgreSQL \cite{PostgreSQL}; \cite{DataModelDataStructureGueting}
and \cite{NetworkGueting} are implemented in the freely available extensible
\secondo{} DBMS \cite{SecondoPlatformPrototypingGueting}.

Although \cite{STAUPelekis} for the DMFS and \cite{DynamicTransportNetworkDing} for
the NCDM provide greater flexibility we decided to use
\cite{DataModelDataStructureGueting} and \cite{NetworkGueting} in our
experiments, because both data models are available in the same DBMS, which is
also the DBMS in which the \bmodb{} \cite{BerlinMODVLDBDuentgen} has bee developed.

The \bmodb{} \cite{BerlinMODVLDBDuentgen} is to the best of our knowledge the only
benchmark testing the capabilities of complete spatio-temporal database systems.
Coming with a well defined data set, and two query sets feasible for DMFS and NCDM.
Other benchmarks for spatio-temporal databases systems provide only well defined
query sets and a database description without any data set like \cite{QueriesTheodoridis}.
Or they come with well defined data generation, workload sets and experiments but
evaluate only the capabilties of indexes for current and near future positions like
\cite{COSTBenchmarkJensen}. Or they focus on time-evolving regional data and associated
index methods like \cite{BenchmarkTzouramanis}.

The focus on index benchmarking in the most benchmarks is dued by the fact that
indexes have a great influence on query run times. Therefore, many spatio-temporal
indexes have been developed in the last ten years. An survey about existing spatio-temporal
indexes can be found in the two parted work \cite{STAMSurvey1} and \cite{STAMSurvey2}.
The most presented spatio-temporal indexes base on the R-Tree \cite{RTreeGuttmann}
and its variants. The R-Trees are used stand alone or in hierarchical combinations.
B-Trees \cite{BTreeBayer} and their variants are also used within spatio-temporal
indexes. \secondo{} comes with implementations of R-Tree, B-Tree, and MON-Tree
\cite{MONTreeAlmeidaGeoinformatica} in Section \ref{sec:implNDM} we give a detailed
description how we used these indexes in our experiments.

The \bmodb{} comes with his own data generator. Like mentioned before other
benchmarks like \cite{QueriesTheodoridis} define only database descriptions.
The users have to generate their own corresponding data.

Therefore, several data generators have been developed. Some of them generate
only unconstrained moving point objects like \cite{OportoSaglio}, or support
only short-term observations like \cite{BrinkhoffsDataGenerator}, or are
unavailable respectively require additional software like \cite{STACTS}.

In the sequel we give short reviews of the \secondo{} DBMS
(Section \ref{sec:secondo}), the both data models we used in our experiments
(Section \ref{sec:bmodbdatamod} and Section \ref{sec:netdatamod}), and the \bmodb{}
(Section \ref{sec:bmodb}).
\subsection{\secondo{} DBMS}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in
\cite{SecondoEnvironmentDieker,SecondoPlatformPrototypingGueting} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce new data types and operations on
these data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the users may publish their extensions
as a \secondo{} plugin such that other users can use these plugins to extend their own
\secondo{} system. They may use the newly provided functionalities or repeat the
published experiments. \secondo{} is freely available on the web \cite{secondoweb}.
It comes with a number of already implemented spatial and spatio-temporal data types
and operations including SPACE (Section \ref{sec:bmodbdatamod}) and NET
(Section \ref{sec:netdatamod}). Furthermore, the \bmodb{}
described in Section \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version 3.0.
\subsection{Data Model of \bmodb{} (SPACE)}
\label{sec:bmodbdatamod}
\cite{STDTModelingQueryingMOinDErwig} presents the basic idea of the DMFS that
is used by the \bmodb{}. The abstract data model for SPACE was pubished in
\cite{RepresentingMovingObjectsGueting} and the discrete data model in
\cite{DataModelDataStructureGueting}. Abstract data models are useful as conceptual
models, but they cannot be implemented, because computers can only use finite sets.
In the sequel description we will mainly focus on the discrete data model.

The type system in \cite{RepresentingMovingObjectsGueting} is defined using the
techniques presented in \cite{SecondOrderSignatureGueting}. The basic idea of
\cite{RepresentingMovingObjectsGueting} is to define type constructors
that create new data types if they are applied to an data type of a given set of
basic data types.

Basic data types are the standard data types integer, real, string and boolean (BASE);
the spatial data types point, points, line, and region (SPATIAL);
and the temporal type instant (TIME).

The carrier sets for all data types in the discrete data model contain $\perp$,
representing an undefined value. This is not mentioned any more in the sequel.

The carrier sets for the BASE data types in the discrete data model are defined
by the corresponding programming language data types \dt{int}, \dt{real}, \dt{bool},
and \dt{string}. The carrier set of a value of the data type \dt{point} is
\dt{real} $\times$ \dt{real}. The two \dt{real} values represent the x,y-coordinates
of the \dt{point} value in the two dimensional plane. The data type \dt{points}
consists of a disjoint set of \dt{point} values. The carrier set for the data type \dt{line}
consists of a finite set of disjoint line segments representing the linear
approximation of the line curve in the two dimensional plane. Semanticaly an
\dt{line} value is the union of the points of all its line segments. A region
is the union of all points covered by the region. The carrier set \dt{region}
is defined to be a finite set of line segments building a polygon representing
the linear approximation of the outer and, if the region contains wholes, inner
borders of the region. The borders are defined to belong to the region.

The carrier set for the TIME data type is given by \dt{real} in the discrete data model.
That means each time \dt{instant} is represented by a corresponding \dt{real} value.

The type constructor \dt{range} converts BASE and TIME data types $\alpha$ into
a type whoes values are finite sets of intervals over $\alpha$. Range types are
used to represent collections of time intervals, or the values taken by a moving
real. Intervals are represented by their start and end point and two flags indicating
if the start respectively end point is part of the interval or not.

The other important type constructor of the abstract data model is \dt{moving}.
In the abstract data model\dt{moving} maps each data type $\alpha$ from BASE and
SPATIAL into an time dependend spatio-temporal moving data type \dt{moving}($\alpha$)
(\dt{m$\alpha$} for short) of kind TEMPORAL. The discrete data model introduces some
additional type constructors to implement the moving type constructor of the
abstract data model. A detailed description of the type constructors is skiped
due to place limitations. We explain the realisation of \dt{moving} at the example
of a \dt{moving}(\dt{point}) (short \dt{mpoint}) object. An \dt{mpoint} value may represent
a car, which changes its position in the plain within time.

An \dt{mpoint} consists of a set of so called \dt{unit}(\dt{point}) values
(\dt{upoint} for short).
Each \dt{upoint} consists of a time interval and two \dt{point} values. The first
\dt{point} value represents the position of the \dt{upoint} at the start of the
time interval and the second \dt{point} value represents the position of the
\dt{upoint} at the end of the time interval. It is assumed that the object represented
by the \dt{upoint} moves on the straight line between these two points with constant
speed within the given time interval. The velocity of the object is given by the
ratio from the distance of the two points and the length of the time interval of
the \dt{upoint}.

All \dt{upoint} values of an \dt{mpoint} must have disjoint time intervals,
because a car cannot be at two different positions at the same time.
The set of \dt{upoint} values is sorted by ascending time intervals.

This spatio-temporal data model of \dt{moving} allows us to compute the position
of an \dt{mpoint} at every time instant within its definition time.
We can also compute the time instant the point passed a
given position assuming the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by an \dt{intime(point)}
(short form \dt{ipoint}). An \dt{ipoint} consists of an time instant and an \dt{point}
value and represents the position of the \dt{mpoint} value at the given time instant.
